---
title: "US Climate Normal"
author: "Dr. Chih-Yu Hung"
date: "2024-05-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse); library(data.table)
```

## To prepare Climate normals

US Climate normal for temperature (degree C), precipitation (mm), potential evapotranspiration (mm) in 1981-2010.
The spatial scale is county and state. 
Original data: monthly data from 1981-2010. 

```{r read data}
# Temp <- fread("Climate Normals/Temperature_perCounty.csv", header = TRUE)
# Pre <- fread("Climate Normals/Precipitation_perCounty.csv", header = TRUE)
Pet <- fread("Climate Normals/PET_perCounty.csv", header = TRUE)

#Name list for missing cities
# unique_Pet <- setdiff(Pet$County_State, Temp$County_State)

```


## Data comparison

The data length of Temp, Pre and Pet are not the same. This is because there are independent cities in Virgina, which are equal to county but small area. There are four cities 
"Fairfax city - Virginia","Falls Church city - Virginia"  "Lexington city - Virginia","Manassas Park city - Virginia
`setdiff` to see the differnences. 

Fairfax city is in Fairfax county, using the same climate
Falls church city is in Fairfax county, using the same climate
Lexington city is in Rockbridge county, using the same climate
Manassas Parck is in Fairfax county, using the same climate


```{r assign climate to the four independent cities}
### No need to do it again. 
#Temp
# FFcounty_T <- Temp[Temp$County_State=="Fairfax County - Virginia",]
# Rbcounty_T <- Temp[Temp$County_State=="Rockbridge County - Virginia",]
# Temp <- rbind(Temp,FFcounty_T, FFcounty_T, Rbcounty_T, FFcounty_T)
# 
# #rename the cities
# n <- length(Temp$County_State)
# Temp$County_State[(n-3):n] <- unique_Pet
# 
# ##Precipitation
# FFcounty_P <- Pre[Pre$County_State=="Fairfax County - Virginia",]
# Rbcounty_P <- Pre[Pre$County_State=="Rockbridge County - Virginia",]
# Pre <- rbind(Pre,FFcounty_P, FFcounty_P, Rbcounty_P, FFcounty_P)
# 
# #rename the cities
# Pre$County_State[(n-3):n] <- unique_Pet
# 
# 
# write.csv(Temp, "Climate Normals/Temperature_perCounty_corrected.csv", row.names = FALSE)
# write.csv(Pre, "Climate Normals/Precipitation_perCounty_corrected.csv", row.names = FALSE)

```


## Temperature data

Data need to be "State", "County", "MAT"
The "MAT" is the average annual temperature during 1981-2010 

```{r Temperature}
Temp <- fread("Climate Normals/Temperature_perCounty_corrected.csv", header = TRUE)
Temp <- Temp %>%
   separate(County_State, into = c("County", "State"), sep = " - ")

# Calculate the average of the 1981-2010
US_Temp <- Temp %>%
  summarise(County = County, 
            State = State, 
            MAT = rowMeans(.[,3:362],na.rm=TRUE))


# Monthly average for 1981-2010
Temp_long <- Temp %>%
  # Reshape so that all P_YYYYMM columns go under "name" and their values under "value"
  pivot_longer(
    cols = starts_with("T_"),     # or use everything() except County/State if needed
    names_to = "name",
    values_to = "val"
  ) %>%
  # Extract year and month from the last 6 digits of column name (e.g., P_198101)
  mutate(
    year  = as.integer(str_sub(name, -6, -3)),   # "1981"
    month = as.integer(str_sub(name, -2, -1))    # "01"
  ) %>%
  # Keep only 1981–2010 data
  filter(year >= 1981 & year <= 2010) %>%
  # Group by County, State, and month (or remove County/State if averaging across all)
  group_by(County, State, month) %>%
  # Compute the mean value (removing NAs)
  summarise(
    mean_Temp = mean(val, na.rm = TRUE),
    .groups  = "drop"
  )


write.csv(Temp_long,"Climate Normals/Monthly_Temp_perCounty_19812010.csv",row.names = FALSE)


```


## Precipitation data

Data need to be "State", "County", "MAP"
The "Pre" is the total annual Precipitation during 1981-2010 

```{r Precipitation}
Pre <- fread("Climate Normals/Precipitation_perCounty_corrected.csv", header = TRUE)
Pre <- Pre %>%
   separate(County_State, into = c("County", "State"), sep = " - ")

# Calculate the average of the 1981-2010
US_Pre <- Pre %>%
  reframe(County = County, 
          State = State,
          MAP = rowSums(.[,3:362])/30)


# Monthly average for 1981-2010
Pre_long <- Pre %>%
  # Reshape so that all P_YYYYMM columns go under "name" and their values under "value"
  pivot_longer(
    cols = starts_with("P_"),     # or use everything() except County/State if needed
    names_to = "name",
    values_to = "val"
  ) %>%
  # Extract year and month from the last 6 digits of column name (e.g., P_198101)
  mutate(
    year  = as.integer(str_sub(name, -6, -3)),   # "1981"
    month = as.integer(str_sub(name, -2, -1))    # "01"
  ) %>%
  # Keep only 1981–2010 data
  filter(year >= 1981 & year <= 2010) %>%
  # Group by County, State, and month (or remove County/State if averaging across all)
  group_by(County, State, month) %>%
  # Compute the mean value (removing NAs)
  summarise(
    mean_Pre = mean(val, na.rm = TRUE),
    .groups  = "drop"
  )

write.csv(Pre_long,"Climate Normals/Monthly_Pre_perCounty_19812010.csv",row.names = FALSE)
```


## Potential evapotranspiration data

Data need to be "State", "County", "MPE"
The "MPE" is the total annual potential evapotranspiration during 1981-2010 

```{r MPE}
Pet <- Pet %>%
   separate(County_State, into = c("County", "State"), sep = " - ")

# Define the number of days in each month for a non-leap year
days_in_month <- c(31, 28, 31, 30, 31, 30, 31, 31, 30, 31, 30, 31)

# Assuming data frame is named 'df' and columns 3 to 362 contain the monthly values
# Convert the relevant columns to a matrix
monthly_values <- as.matrix(Pet[, 3:362])

# Create a vector repeating 'days_in_month' 30 times to match the 360 months
days_vector <- rep(days_in_month, times = 30)

# Multiply each column by the corresponding days in month
adjusted_values <- sweep(monthly_values, 2, days_vector, "*")

# Replace the original monthly values with the adjusted values in the data frame
Pet[, 3:362] <- adjusted_values

# Calculate the average of the 1981-2010
US_Pet <- Pet %>%
  reframe(County = County, 
          State = State,
          MPE = rowSums(.[,3:362])/30)


#Monthly average for PET in 1981-2010
Pet_long <- Pet %>%
  # Reshape so that all P_YYYYMM columns go under "name" and their values under "value"
  pivot_longer(
    cols = starts_with("PET_"),     # or use everything() except County/State if needed
    names_to = "name",
    values_to = "val"
  ) %>%
  # Extract year and month from the last 6 digits of column name (e.g., P_198101)
  mutate(
    year  = as.integer(str_sub(name, -6, -3)),   # "1981"
    month = as.integer(str_sub(name, -2, -1))    # "01"
  ) %>%
  # Keep only 1981–2010 data
  filter(year >= 1981 & year <= 2010) %>%
  # Group by County, State, and month (or remove County/State if averaging across all)
  group_by(County, State, month) %>%
  # Compute the mean value (removing NAs)
  summarise(
    mean_Pet = mean(val, na.rm = TRUE),
    .groups  = "drop"
  )


write.csv(Pet_long,"Climate Normals/Monthly_Pet_perCounty_19812010.csv",row.names = FALSE)

```
