---
title: "1.3 Growing season"
author: "Dr. Chih-Yu Hung"
date: "2025-03-4"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(data.table)

Temp <- read_csv("Climate Normals/Monthly_Temp_perCounty_19812010.csv")

```


## Determine if we are going to apply coef of non-growing season 

The reason why we need to apply non-growing season coef is the spring freeze-thaw cycles (FTC). The are hot moments cause soil N2O emissions in temperate regions. The intensity of freeze also influence the magnitude of soil N2O emissions. 

We will calculate Cumulative freeze degree day (CFD) to determine if we want to apply the non-growing season factor. (Wagner-Riddle et. al., 2017)

The f(x) = 1.98(1−e(−0.00724x)), We use  XXX method and then determine when CFD more than 96 it has non-growing season 

Ref.
DOI: 10.1038/ngeo2907
DOI: 10.1016/j.agee.2021.107620
DOI: 10.1139/cjss-2022-0101
DOI: 10.1016/j.apsoil.2020.103786



```{r NGS coef}

# Create a lookup for days in each month, ignoring leap years
days_lookup <- c(31, 28, 31, 30, 31, 30, 31, 31, 30, 31, 30, 31)

NGS <- Temp %>%
  # Attach the day count based on month
  mutate(days = days_lookup[month]) %>%
  # Calculate contribution only when mean_Temp < 0
  mutate(temp_contrib = if_else(mean_Temp < 0, mean_Temp * days, 0)) %>%
  # Sum the contributions for each County-State combination
  group_by(County, State) %>%
  summarise(cumulative_temp = sum(temp_contrib), .groups = "drop") %>%
  # Create the indicator column
  mutate(NGS = if_else(cumulative_temp < -96, "Y", "N"))


NGS <- NGS %>%
  mutate(County = str_replace(County, " County$", "")) %>%
  mutate(State = toupper(State))



write.csv(NGS,"Outputs/1.3_Non-growing season_30y.csv", row.names = FALSE)
```

