---
title: "Quick data_US"
author: "Dr. Chih-Yu Hung"
date: "2024-11-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
source("setup.R")
#2017 data
#US_census2017 <- fread("qs.census2017.txt/qs.census2017.txt")
#save(US_census2017,file = "Inputs/US_census2017.RData")
#load("Inputs/US_census2017.RData")

#2022 data
#US_census2022 <- fread("qs.census2022.txt/qs.census2022.txt")
#save(US_census2022,file = "Inputs/US_census2022.RData")
#load("Inputs/US_census2022.RData")

#US_census_crop <- fread("qs.crops_20241107.txt/qs.crops_20241107.txt") #3-4 minutes to read
#save(US_census_crop,file = "Inputs/US_census_crop.RData")
#load("Inputs/US_census_crop.RData") #still need 1-2 minutes to load


```

## Filter the data

Exclude the states such as AK and HI. and keep 2022 and target crop data only

### Operate area
To know the total area that could have tillage operation, filters were applied
1. PRODN_PRACTICE_DESC = "ALL PRODUCTION PRACTICES"
2. STATISTICCAT_DEC = "AREA HARVESTED"
3. UNIT_DESC = "ACRES"
4. DOMAIN_DESC = "AREA OPERATED"
5. AGG_LEVEL_DESC = "STATE"  (to get state level data)

After the filter, sum of "VALUE" is the total area (acre) for the crop in the state.

```{r obtain data for target crops}
# data_2022 <- US_census2022 %>%
#   filter(!(STATE_ALPHA %in% c("AK","GU","AS","PR","VI","MP","","HI",""))) %>% #exclude territories, HI and AK
#   filter(COMMODITY_DESC %in% toupper(target_crops))
#save(data_2022, file = "Inputs/data_crop_2022.RData")

# data_2017 <- US_census2017 %>%
#   filter(!(STATE_ALPHA %in% c("AK","GU","AS","PR","VI","MP","","HI",""))) %>% #exclude territories, HI and AK
#   filter(COMMODITY_DESC %in% toupper(target_crops))
#save(data_2017, file = "Inputs/data_crop_2017.RData")

load("Inputs/data_crop_2022.RData")
load("Inputs/data_crop_2017.RData")

#To filter the area data in 2022 and 2017
data_area_2022 <- data_2022 %>%
  filter(PRODN_PRACTICE_DESC == "ALL PRODUCTION PRACTICES") %>%
  filter(STATISTICCAT_DESC == "AREA HARVESTED") %>%
  filter(UNIT_DESC == "ACRES") %>%
  filter(DOMAIN_DESC == "AREA OPERATED") %>%
  filter(AGG_LEVEL_DESC == "STATE") %>%
  mutate(Area_tot = ifelse(grepl("^\\(", VALUE), 0, VALUE)) %>%
  mutate(Area_tot = as.numeric(gsub(",", "", Area_tot))) %>%   
  group_by(COMMODITY_DESC, SHORT_DESC) %>%
  summarise(Area_tot = sum(Area_tot))
  

data_area_2017 <- data_2017 %>%
  filter(PRODN_PRACTICE_DESC == "ALL PRODUCTION PRACTICES") %>%
  filter(STATISTICCAT_DESC == "AREA HARVESTED") %>%
  filter(UNIT_DESC == "ACRES") %>%
  filter(DOMAIN_DESC == "AREA OPERATED") %>%
  filter(AGG_LEVEL_DESC == "STATE") %>%
  mutate(Area_tot = ifelse(grepl("^\\(", VALUE), 0, VALUE)) %>%
  mutate(Area_tot = as.numeric(gsub(",", "", Area_tot))) %>%   
  group_by(COMMODITY_DESC, SHORT_DESC) %>%
  summarise(Area_tot = sum(Area_tot))


#To filter the production data in 2022 and 2017
data_prod_2022 <- data_2022 %>%
  filter(PRODN_PRACTICE_DESC == "ALL PRODUCTION PRACTICES") %>%
  filter(STATISTICCAT_DESC == "PRODUCTION") %>%
  filter(UNIT_DESC %in% c("BU","TONS","LB")) %>% #they use dif units for grain and silage, or other
  filter(DOMAIN_DESC == "AREA OPERATED") %>%
  filter(AGG_LEVEL_DESC == "STATE") %>%
  mutate(Prod_tot = ifelse(grepl("^\\(", VALUE), 0, VALUE)) %>%
  mutate(Prod_tot = as.numeric(gsub(",", "", Prod_tot))) %>%   
  group_by(COMMODITY_DESC, SHORT_DESC) %>%
  summarise(Prod_tot = sum(Prod_tot))


data_prod_2017 <- data_2017 %>%
  filter(PRODN_PRACTICE_DESC == "ALL PRODUCTION PRACTICES") %>%
  filter(STATISTICCAT_DESC == "PRODUCTION") %>%
  filter(UNIT_DESC %in% c("BU","TONS","LB")) %>% #they use dif units for grain and silage, or other
  filter(DOMAIN_DESC == "AREA OPERATED") %>%
  filter(AGG_LEVEL_DESC == "STATE") %>%
  mutate(Prod_tot = ifelse(grepl("^\\(", VALUE), 0, VALUE)) %>%
  mutate(Prod_tot = as.numeric(gsub(",", "", Prod_tot))) %>%   
  group_by(COMMODITY_DESC, SHORT_DESC) %>%
  summarise(Prod_tot = sum(Prod_tot))

```



