---
title: "2.1 Soil texture"
author: "Dr. Chih-Yu Hung"
date: "2025-02-26"
output: html_document
---

## Soil texture data from SSURGO

The Soil Survey Geographic Database (SSURGO) collected by the National Cooperative Soil Survey over the course of a century. 
(https://www.nrcs.usda.gov/resources/data-and-reports/soil-survey-geographic-database-ssurgo)

They provide an R package, soilDB, to obtain data from SSURGO

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
source("setup.R")

```

## Determine Soil texture class

The soil texture class are coarse, medium and fine in our N2O emissions estimate. 

```{r Soil texture class}
#Classifiy based on USDA soil texture. Medium and Coarse have 0.49 ratio and Fine have 2.55 ratio
classify_soil_texture <- function(clay, sand) {
  if (clay >= 35) {
    return("Fine")
  } else if (clay >= 27.5 & sand < 45) {
    return("Fine")
  } else {
    return("Med/Coarse")
  }
}
```


## Soil sand, silt, clay content from soil texture

Every state has multiple mukey, which is the least survey unit in SSURGO. Each mukey has its unique sand, silt, clay content. I'm going to use this to determine their soil texture and based on their area to determine the percentage of Fine soil and Med/Coarse soil in the State. 


```{r Loop all state for soil texture}
# Loop through each state abbreviation in state_crosswalk
# Initialize an empty dataframe to store results
all_states_texture_summary <- data.frame()

# Loop through each state abbreviation in state_crosswalk
for (state_code in state_crosswalk$abbr) {
  
  # Fetch soil texture data (sand, silt, clay) by mukey
  soil_texture <- get_SDA_property(
    property = c("sandtotal_r", "silttotal_r", "claytotal_r"),
    method = "Weighted Average",
    WHERE = paste0("areasymbol LIKE '", state_code, "%'"),
    top_depth = 0,
    bottom_depth = 30
  ) %>% select(mukey, sandtotal_r, silttotal_r, claytotal_r)

  # Fetch area data using mukey from get_mapunit_from_SDA()
  area_data <- get_mapunit_from_SDA(
    WHERE = paste0("areasymbol LIKE '", state_code, "%'")
  ) %>% select(mukey, muacres)

  # Merge both datasets by mukey
  soil_texture_area <- left_join(soil_texture, area_data, by = "mukey")

  # Remove rows with NA values
  soil_texture_area_clean <- soil_texture_area %>%
    na.omit()

  # Skip this iteration if no valid data is found for the state
  if (nrow(soil_texture_area_clean) == 0) next

  # Classify soil texture for each mukey
  soil_texture_area_clean <- soil_texture_area_clean %>%
    mutate(Soil_Texture = mapply(classify_soil_texture, claytotal_r, sandtotal_r))

  # Calculate total area
  total_area <- sum(soil_texture_area_clean$muacres, na.rm = TRUE)

  # Compute percentage of each soil texture type
  # We don't need to separate coarse and medium because they both use 0.49.
  fine_area <- sum(soil_texture_area_clean$muacres[soil_texture_area_clean$Soil_Texture == "Fine"], na.rm = TRUE)
  med_coarse_area <- sum(soil_texture_area_clean$muacres[soil_texture_area_clean$Soil_Texture == "Med/Coarse"], na.rm = TRUE)

  # Calculate percentage of each soil texture category
  fine_percentage <- (fine_area / total_area) * 100
  med_coarse_percentage <- (med_coarse_area / total_area) * 100

  # Create a summary dataframe for the state
  state_texture_summary <- data.frame(
    State = state_code,
    Total_Area_Acres = total_area,
    Frac_Fine = fine_percentage,
    Frac_Medium = med_coarse_percentage/2,
    Frac_Coarse = med_coarse_percentage/2
  )

  # Append to the final results dataframe
  all_states_texture_summary <- bind_rows(all_states_texture_summary, state_texture_summary)
}

all_states_texture_summary <- all_states_texture_summary %>%
  left_join(state_crosswalk, by = c("State" = "abbr")) %>%
  select(-State) %>%
  rename(State = full) %>%
  mutate(State = toupper(State))


#write.csv(all_states_texture_summary,"Outputs/2.1_Soil texture.csv", row.names = FALSE)
```

