---
title: "US Inorganic fertilization"
author: "Dr. Chih-Yu Hung"
date: "2025-01-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(data.table)
```


## Fertilization County level and crop specific

I incorporate my estimate of total fertilizer use (crop specific and county level) and calibrate it with US NIR county level estimate. The proportion of fertilizer type will be applied for volatilization estimate.


```{r fertilization county level}
Fertilizer_County_Crop <- read.csv("Outputs/3.4_Fertilizer_Use_County_Crop.csv")
Fert_Form <- read.csv("Outputs/3.2_Fertilizer_form.csv")
Fert_Calibration <- read.csv("Outputs/3.6_Fertilizer_Calibration.csv") %>%
na.omit(.) # Some rows with NA as there is no value in CNTY_FERT_1987-2012


#Adjust the fertilizer use for crops 
Fert_County_Crop <- Fertilizer_County_Crop %>%
left_join(Fert_Calibration, by = c("STATE", "COUNTY", "YEAR")) %>%
mutate(Fert_LCA_kg = FERT_kg/RATIO) %>%
select(STATE,COUNTY,YEAR,CROP, CLASS, DOUBLE,Fert_LCA_kg, AREA)

#write.csv(Fert_County_Crop,"Outputs/3.0_Fertilizer_County_Crop_LCA.csv",row.names = FALSE)
```



##NOTE IMPOTANT!!!!!
### BEfore I do anything remember that this fertilzation amount is the total of all crops. I only want some crops. The sum of fertilzer of the interested crop must be less than the total fertilizer in this list. 
(CY, 2025 Jan 09)

```{r retrieve survey data for fertilzation rate}
#  US_environment <- fread("qs.environmental_20250114.txt/qs.environmental_20250114.txt")
# # 
# Fert_rate <- US_environment %>%
#   filter(SOURCE_DESC == "SURVEY") %>% #keep CENSUS data only
#   filter(!(STATE_ALPHA %in% c("AK","GU","AS","PR","VI","MP","","HI",""))) %>% #exclude territories, HI and AK
#   filter(YEAR %in% c(2018:2022)) %>%
#   filter(COMMODITY_DESC %in% c("CORN","SORGHUM","GRAIN",
#                                "WHEAT","SOYBEANS","OATS","BARLEY",
#                                "SWEET CONR","PEAS","RYE","CANOLA",
#                                "TRITICALE","CAMELINA")) %>%
#   filter(DOMAINCAT_DESC == "FERTILIZER: (NITROGEN)") %>%
#   filter(UNIT_DESC == "LB / ACRE / YEAR, AVG") %>%
#   select(COMMODITY_DESC,CLASS_DESC,STATE_ALPHA,STATE_NAME,YEAR,VALUE)
# save(Fert_rate, file = "Inputs/Fertrate_2018-22.RData")

load("Inputs/Fertrate_2018-22.RData") #unit lb /acre

Fert_rate <- Fert_rate %>%
  filter(VALUE != "(D)") %>%
  mutate(VALUE = as.numeric(VALUE)) %>%
  mutate(CROP = COMMODITY_DESC ) %>%
  mutate(CLASS = CLASS_DESC) %>%
  group_by(CROP,CLASS,STATE_NAME)%>%
  summarise(RATE = mean(VALUE))

```


