---
title: "3.1 Fertilization rate"
author: "Dr. Chih-Yu Hung"
date: "2025-01-30"
output: html_document
---

## Fertilization rate

The fertilization rates are from survey data and regional fertilzation recommendation. 

About corn, soybeans, and wheat fertilization rate. There is an Excel file from USDA ERS (https://www.ers.usda.gov/data-products/fertilizer-use-and-price), which only include fertilization rate until 2018. This is the data many research cite. I use the survey data from 2018- 2022 and take average for the fertilization rate.


```{r retrieve survey data for fertilzation rate}
#  US_environment <- fread("qs.environmental_20250114.txt/qs.environmental_20250114.txt")
# # 
# Fert_rate <- US_environment %>%
#   filter(SOURCE_DESC == "SURVEY") %>% #keep CENSUS data only
#   filter(!(STATE_ALPHA %in% c("AK","GU","AS","PR","VI","MP","","HI",""))) %>% #exclude territories, HI and AK
#   filter(YEAR %in% c(2018:2022)) %>%
#   filter(COMMODITY_DESC %in% c("CORN","SORGHUM","GRAIN",
#                                "WHEAT","SOYBEANS","OATS","BARLEY",
#                                "SWEET CONR","PEAS","RYE","CANOLA",
#                                "TRITICALE","CAMELINA")) %>%
#   filter(DOMAINCAT_DESC == "FERTILIZER: (NITROGEN)") %>%
#   filter(UNIT_DESC == "LB / ACRE / YEAR, AVG") %>%
#   select(COMMODITY_DESC,CLASS_DESC,STATE_ALPHA,STATE_NAME,YEAR,VALUE)
# save(Fert_rate, file = "Inputs/Fertrate_2018-22.RData")

load("Inputs/Fertrate_2018-22.RData") #unit lb /acre

Fert_rate <- Fert_rate %>%
  filter(VALUE != "(D)") %>%
  mutate(VALUE = as.numeric(VALUE)) %>%
  mutate(CROP = COMMODITY_DESC ) %>%
  mutate(CLASS = CLASS_DESC) %>%
  group_by(CROP,CLASS,STATE_NAME)%>%
  summarise(RATE = mean(VALUE))

```

## Complete the data table

There are crops and states missing because they are not included in the survey. 
Crop: camelina, canola, rye (no spring or fall in the US) triticale


```{r complete data table}
missing_crops <- c("CAMELINA", "CANOLA", "RYE", "TRITICALE")

states <- toupper(c(
  "Alabama", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", 
  "Florida", "Georgia", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", 
  "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", 
  "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", 
  "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", 
  "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", 
  "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", 
  "Wisconsin", "Wyoming"
))

# Step 1: Ensure all states and crops are included for existing crops and their classes
# Create a full grid of all combinations of crops, classes, and states from the original data
all_combinations <- Fert_rate %>%
  select(CROP, CLASS) %>%
  distinct() %>%
  crossing(STATE_NAME = states) # Create combinations with all states

# Merge the original data with the full grid
completed_existing_data <- all_combinations %>%
  left_join(Fert_rate, by = c("CROP", "CLASS", "STATE_NAME"))


# Step 2: Add missing crops for all states
missing_data <- expand.grid(
  CROP = missing_crops,
  STATE_NAME = states,
  CLASS = "ALL CLASSES",
  RATE = NA, # Assuming RATE is missing for new rows
  stringsAsFactors = FALSE
)

# Step 3: Combine the completed data
Fert_rate_expanded <- bind_rows(completed_existing_data, missing_data) %>%
  distinct() # Remove duplicates

```

## Fertilizer rates for crops
We don't have fertilizer rate for every states.  Find the major production states and their fertilizer rates. and then apply their rates to the states in the same Farm Production Region.  

1. For corn, take Grain corn only (Silage production is only about 1% of grain)
2. For sorghum, take Grain only
3. For Peas, take "Dry edible" which is the total of others
4. For Wheat, we need to seprate to SPRING, (EXCL DURUM), SPRING, DURUM and WINTER

```{r prodcution percentage calculation}
load("Inputs/US_census2022.RData")

# Initialize an empty list to store production tables
production_tables <- list()

Production_percentage <- function(data, crop_class) {
  # Extract crop and class information
  crop_info <- strsplit(crop_class, "_")[[1]]
  crop <- crop_info[1]
  class <- ifelse(length(crop_info) > 1, crop_info[2], NA)  # Class is NA for non-WHEAT crops

  # Filter the dataset for the specified crop and relevant conditions
  filtered_data <- data %>%
    filter(COMMODITY_DESC == crop) %>%
    filter((COMMODITY_DESC != "CORN") | (COMMODITY_DESC == "CORN" & UTIL_PRACTICE_DESC == "GRAIN")) %>%
    filter((COMMODITY_DESC != "SORGHUM") | (COMMODITY_DESC == "SORGHUM" & UTIL_PRACTICE_DESC == "GRAIN")) %>%
    filter((COMMODITY_DESC != "PEAS") | (COMMODITY_DESC == "PEAS" & CLASS_DESC == "DRY EDIBLE")) %>%
    filter(STATISTICCAT_DESC == "PRODUCTION") %>%
    filter(AGG_LEVEL_DESC %in% c("NATIONAL", "STATE")) %>%
    filter(DOMAIN_DESC == "TOTAL") %>%
    mutate(VALUE = as.numeric(gsub(",", "", VALUE))) # Remove commas

  # Apply additional filtering for WHEAT classes
  if (!is.na(class)) {
    filtered_data <- filtered_data %>%
      filter(CLASS_DESC == class)
  }

  # Extract UNIT_DESC from filtered_data
  unit <- unique(filtered_data$UNIT_DESC)
  if (length(unit) > 1) {
    warning("Multiple units found. Using the first one.")
    unit <- unit[1]
  }

  # Calculate US total production
  us_total <- filtered_data %>%
    filter(LOCATION_DESC == "US TOTAL") %>%
    summarize(Total_Production = sum(VALUE, na.rm = TRUE)) %>%
    pull(Total_Production)

  # Calculate state production percentages
  state_production <- filtered_data %>%
    filter(LOCATION_DESC != "US TOTAL") %>%
    group_by(LOCATION_DESC) %>%
    summarize(State_Production = sum(VALUE, na.rm = TRUE), .groups = "drop") %>%
    mutate(
      Production_Percentage = round((State_Production / us_total) * 100, 1),
      CROP = crop_class,
      UNIT = unit  # Add unit to the output
    ) %>%
    arrange(desc(Production_Percentage))

  # Save the table to the list
  production_tables[[crop_class]] <<- state_production
}

# Redefine unique_crops to include CLASS_DESC for WHEAT
unique_crops <- Fert_rate_expanded %>%
  mutate(CROP_CLASS = ifelse(CROP == "WHEAT", paste(CROP, CLASS, sep = "_"), CROP)) %>%
  distinct(CROP_CLASS) %>%
  pull(CROP_CLASS)

# Calculate production percentages for all crop-class combinations
for (crop_class in unique_crops) {
  Production_percentage(US_census2022, crop_class)
}

# Combine all production tables into one data frame
combined_production <- do.call(rbind, lapply(production_tables, function(df) {
  df %>%
    mutate(LOCATION_DESC = as.character(LOCATION_DESC)) # Ensure consistent column types
}))

# Join the combined production data back to Fert_rate_expanded
final_result <- Fert_rate_expanded %>%
  mutate(CROP_CLASS = ifelse(CROP == "WHEAT", paste(CROP, CLASS, sep = "_"), CROP)) %>%
  left_join(combined_production, by = c("CROP_CLASS" = "CROP", "STATE_NAME" = "LOCATION_DESC"))

# Clean up unnecessary columns, rename
final_result <- final_result %>%
  select(CROP,  CLASS, STATE_NAME,State_Production, Production_Percentage, RATE, UNIT) %>%
  rename(STATE = STATE_NAME,
         PRODUCTION = State_Production,
         PERCENTAGE = Production_Percentage) %>%
  # Replace NA in PRODUCTION and PERCENTAGE with 0
  mutate(
    PRODUCTION = replace_na(PRODUCTION, 0),
    PERCENTAGE = replace_na(PERCENTAGE, 0)
  ) %>%
  # Replace NA in UNIT with the unit from the same CROP and CLASS
  group_by(CROP, CLASS) %>%
  mutate(
    UNIT = if_else(is.na(UNIT), first(na.omit(UNIT)), UNIT)
  ) %>%
  ungroup()

#Unit conversion
# Conversion factors
conversion_factors <- list(
  BU = 30.027682980816,  # Bushels
  CWT = 45.36,           # Hundredweight
  LB = 0.4536            # Pounds
)

# Convert UNIT to kg and adjust PRODUCTION values
Crop_Fert_Prod <- final_result %>%
  mutate(
    PRODUCTION_KG = case_when(
      UNIT == "BU" ~ PRODUCTION * conversion_factors$BU,
      UNIT == "CWT" ~ PRODUCTION * conversion_factors$CWT,
      UNIT == "LB" ~ PRODUCTION * conversion_factors$LB,
      TRUE ~ PRODUCTION  # Keep as is if UNIT does not match
    ),
    UNIT = "KG")%>%   # Update unit to KG 
    select(CROP,  CLASS, STATE, PRODUCTION_KG, PERCENTAGE, RATE )


```




## USing the USDA Farm Production Regions

USDA Farm production regions are useful to estimate fertilizer rate where is missing. Each region defines a geographic area in which crops and cropping practices are similiar. (Cooter et al., 2012)

NOTE: I changed Washington State from Pacific to Mountain, since most crop production in WA is in the valleys of Rocky mountains

Cooter et al., 2012. doi:10.5194/bg-9-4023-2012

```{r Farm production regions}
FPR <- read.csv("Fertilizer/Farm Production Regions.csv") %>%
  mutate(STATE_NAME = toupper(State)) %>%
  select(STATE_NAME, Region) %>%
  rename(STATE = STATE_NAME,
         REGION = Region)
#Change WA to Mountain
FPR <- FPR %>%
  mutate(
    REGION = case_when(
      STATE == "WASHINGTON" ~ "Mountain",
      TRUE ~REGION
  ))


Crop_Fert_Prod_FPR <- Crop_Fert_Prod %>%
  left_join(FPR, by = "STATE")

#write.csv(Crop_Fert_Prod_FPR,"Outputs/Fertilizer rate.csv", row.names = FALSE) #to check my result
```



##Estimate the Fertilizer rate based on weighted production value and reported fertilizer rate

```{r Estimate Fertilizer rate}
# Define a function to calculate the weighted average and fill missing values
fill_missing_rate <- function(data) {
  data %>%
    group_by(CROP, CLASS, REGION) %>% # Group by CROP, CLASS, and REGION
    mutate(
      RATE = ifelse(
        is.na(RATE),
        # Calculate weighted average for non-NA RATE values within the group
        round(sum(RATE * PERCENTAGE, na.rm = TRUE) / sum(PERCENTAGE[!is.na(RATE)], na.rm = TRUE),1),
        RATE
      )
    ) %>%
    # Ensure rows with no valid RATE in the group remain NA
    mutate(RATE = ifelse(is.nan(RATE), NA, RATE)) %>%
    ungroup()
}

# Apply the function to the dataset
Crop_Fert_Prod_FPR <- fill_missing_rate(Crop_Fert_Prod_FPR)


# Generate summary for missing RATE
missing_rate_summary <- Crop_Fert_Prod_FPR %>%
  filter(is.na(RATE)) %>% # Filter rows with missing RATE
  group_by(CROP, CLASS, REGION) %>% # Group by CROP, CLASS, and REGION
  summarize(
    Missing_Count = n(), # Count missing RATE values
    Total_Percentage = sum(PERCENTAGE, na.rm = TRUE) # Sum PERCENTAGE
  ) %>%
  ungroup() # Ungroup the data

```


## Find fertilzer rate for missing values. 

I need references to know fertilizer recommendations for each crop in each regions. Most of them from land grant university or USDA. 
All references can be found in "US data/Fertilizer/Reference"

1. Canola / Northern Plain: North Dakota 60 lb/a = 67.25 N kg/ha
2. Camelina/ Mountain: Montana 60 N kg/ha
3. Peas/ Mountain and Northern Plain: Montana 10 lb/a = 11.21 N kg/ha
4. Triticale / Pacific: California 250 lb/a = 280.2 N kg/ha(irrigated), 30 lb/a = 33.63 N kg/ha (dryland)
               44% land in CA is irrgated. so I used 250 * 0.44 + 30 * 0.56  = 126.8 lb/a          
5. Triticale /Southern and Northern plain: 70 lb/a = 78.46 N kg/ha, should be applicable to Great plain area.
6. Rye/ Northern plain: 60 lb/a = 67.25 N kg/ha
7. Rye/ Lake: 50lb/a = 56 N kg/ha
8. Rye and Triticale/ Northeast: Vermont 60 lb/a = 67.25 N kg/ha
9. Rye/ Southern plains: Oklahoma: 180 lb/a = 201.75 N kg/ha
10. Rye/ Appalachia: North Carolina: 90 lb/a = 100.88 N kg/ha
11. Rye/ Southeast: Georgia: 30+50+50 = 130 lb/a = 144.72 M kg/ha

```{r merge fertilzation rate}
Fert_Ref <- read.csv("Fertilizer/Fertilization_Rate_Reference.csv", header = T)

#Merge the fert rate refernce
Crop_Fert_Prod_FPR <- Crop_Fert_Prod_FPR %>%
  left_join(Fert_Ref, by = c("CROP", "REGION")) %>%
  mutate(RATE = coalesce(RATE.y, RATE.x)) %>%
  select(-RATE.x, -RATE.y)

#Estimate fert rate based on weighted national average fertilzation rate
Crop_Fert_Prod_FPR<- Crop_Fert_Prod_FPR %>%
  group_by(CROP, CLASS) %>%
  mutate(
    weighted_avg_RATE = round(sum(RATE * PERCENTAGE, na.rm = TRUE) / sum(PERCENTAGE[!is.na(RATE)], na.rm = TRUE),1),
    RATE = ifelse(is.na(RATE), weighted_avg_RATE, RATE)
  ) %>%
  ungroup() %>%
  select(-weighted_avg_RATE) %>% # Remove the temporary column
  mutate(`RATE(kg/ha)` = round(RATE*1.12085,2))


#write the results
write.csv(Crop_Fert_Prod_FPR, "Outputs/Fertilization_Rate_State_Crop.csv")

```

