---
title: "Fertilizer types"
author: "Dr. Chih-Yu Hung"
date: "2025-01-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
```

## To obtain fertilizer types in different regions in the US. 

The methodologies to estimate fertilizer types (% for various types) are developed by Cao et al. 
Although it's not crop sepcific and only estimated the proportion to 2015 (and before). 

Only corn has the exception. 
Nitrogen fertilizer types of corn is estimated ny Cooter et al., 2012. 


Ref. 
Cao, P., Lu, C., & Yu, Z. (2018).
Historical nitrogen fertilizer use in agricultural ecosystems of the contiguous United States during 1850–2015: application rate, timing, and fertilizer types. Earth System Science Data .https://doi.org/10.5194/essd-10-969-2018
Cooter, E.J, Bash, J.O. Benson, V. & Ran L. (2012).
Linking agricutural crop management and air quality models for regional to national-scale nitrogen assessments. Biogeosciens



```{r fertilizer form for corn}
Fertilizer_form_corn <- read_excel("Fertilizer/Fertilizer form corn.xlsx", 
    sheet = "Fertilizer form") 
#The percentages don't add up to 100% becasue it has manure. NEED to convert to Synthetic N fertilizer only

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
