---
title: "Fertilizer types"
author: "Dr. Chih-Yu Hung"
date: "2025-01-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(magick)
```

## To obtain fertilizer types in different regions in the US. 

The methodologies to estimate fertilizer types (% for various types) are developed by Cao et al. 
Although it's not crop sepcific and only estimated the proportion to 2015 (and before). 

Only corn has the exception. 
Nitrogen fertilizer types of corn is estimated ny Cooter et al., 2012. 


Ref. 
Cao, P., Lu, C., & Yu, Z. (2018).
Historical nitrogen fertilizer use in agricultural ecosystems of the contiguous United States during 1850–2015: application rate, timing, and fertilizer types. Earth System Science Data .https://doi.org/10.5194/essd-10-969-2018
Cooter, E.J, Bash, J.O. Benson, V. & Ran L. (2012).
Linking agricutural crop management and air quality models for regional to national-scale nitrogen assessments. Biogeosciens



```{r fertilizer form for corn}
Fertilizer_form_corn <- read_excel("Fertilizer/Fertilizer form corn.xlsx", 
    sheet = "Fertilizer form") 
#The percentages don't add up to 100% becasue it has manure. NEED to convert to Synthetic N fertilizer only

Fertilizer_form_corn <- Fertilizer_form_corn %>%
  group_by(REGION) %>%
  mutate(PERCENTAGE = round((PERCENTAGE/ sum(PERCENTAGE)) *100,2)) %>%
  ungroup()

```

## Extract the figure in Cao and calculate the proportion. 

The figure is save to "Fertilizer"


```{r fertilizer form by Cao}

# Load image
image_path <- "Fertilizer/Fertilizer proportion_Cao/Proportion_2015.png"
img <- image_read(image_path)

# Convert image to raster data
img_data <- as.data.frame(as.raster(img))

# Get image dimensions
img_width <- image_info(img)$width
img_height <- image_info(img)$height

# Define the number of bars (regions)
num_bars <- 7
bar_width <- img_width / num_bars  # Approximate width of each bar

# Define region names (manually update if necessary)
region_names <- c("NGP", "MW", "NW", "NE", "SW", "SE", "SGP")

# Add REGION information based on pixel position
img_data <- img_data %>%
  mutate(pixel_position = rep(1:img_width, each = img_height)) %>%  # Simulate x-coordinates
  mutate(REGION = cut(pixel_position, breaks = seq(0, img_width, length.out = num_bars + 1), labels = region_names, include.lowest = TRUE)) %>%
  select(-pixel_position)  # Remove unnecessary column


# Define legend colors, This is from the legend png
legend_colors <- list(
  AnA = rgb(0, 0, 0, maxColorValue = 255),
  AqA = rgb(255, 0, 0, maxColorValue = 255),
  AN = rgb(0, 0, 255, maxColorValue = 255),
  AS = rgb(255, 0, 255, maxColorValue = 255),
  NS = rgb(0, 132, 0, maxColorValue = 255),
  SN = rgb(0, 0, 132, maxColorValue = 255),
  UR = rgb(132, 0, 255, maxColorValue = 255),
  APs = rgb(132, 0, 132, maxColorValue = 255),
  CN = rgb(132, 0, 0, maxColorValue = 255),
  DAP = rgb(132, 132, 0, maxColorValue = 255),
  MAP = rgb(0, 132, 132, maxColorValue = 255)
)

match_colors <- function(pixel_color, legend_colors) {
  color_distances <- sapply(legend_colors, function(color) sum(abs(col2rgb(pixel_color) - col2rgb(color))))
  return(names(which.min(color_distances)))  # Get the closest matching color
}

# Process image data and classify pixels
color_counts <- img_data %>%
  rename(pixel_color = value) %>%
  group_by(pixel_color) %>%
  summarise(count = n(), .groups = 'drop') %>%
  mutate(fertilizer_type = sapply(pixel_color, match_colors, legend_colors)) %>%
  group_by(fertilizer_type) %>%
  summarise(proportion = sum(count) / sum(color_counts$count) * 100, .groups = 'drop')  # Convert to percentage

# Display table
print(color_counts)




```
