---
title: "3.6 Fertilizer Calibration"
author: "Dr. Chih-Yu Hung"
date: "2025-02-11"
output: html_document
---

## Calibrate my estimate of fertilizer with US NIR's estimate

It's possible I overestimate/underestimate total fertilizer usage in a county. 

Overestimate:
1. CDL could overestimate crop area, causing overestimate fertilizer use
2. I could overestimate fertilization rate as I used local recommendation. 
The actual fertilzation is usally adjusted based on soil test result. 
3. Manure and other organic fertilzer are not fully considered in this analysis

Underestimate:
1. CDL could underestimate area of crop with higher fertilizer need. 
2. Underestimate fertilization rate is possbile as I used the average rate. 


```{r Calibration}
library(dplyr)
library(tidyr)
library(stringr)
library(readr)

# read the CSV files
Fert_County_1822 <- read.csv("Outputs/3.5_Fertilizer_County_1822.csv")
Fertilizer_County_Crop <- read.csv("Outputs/3.4_Fertilizer_Use_County_Crop.csv")
state_crosswalk <- read.csv("Inputs/state_crosswalk.csv")



# 1) Pivot Fert_County_1822 from wide to long and create a YEAR column
Fert_County_1822_long <- Fert_County_1822 %>%
  pivot_longer(
    cols = starts_with("farmN"),
    names_to = "farm_year",
    values_to = "farm_val"
  ) %>%
  mutate(
    # Remove the "farmN" prefix, convert to numeric
    YEAR = as.numeric(str_remove(farm_year, "farmN"))
  )

# 2) Convert state abbreviations (in the "State" column) to full names
Fert_County_1822_long <- Fert_County_1822_long %>%
  left_join(state_crosswalk, by = c("State" = "abbr")) %>%
  mutate(STATE = toupper(full)) %>%
  select(-full)  # keep it clean, removing unneeded columns

# 3) Join both tables by STATE, County, and YEAR
Fert_Calibration <- Fert_County_1822_long %>%
  inner_join(
    Fertilizer_County_Crop,
    by = c("STATE", "County" = "COUNTY", "YEAR")
  )

# Now Fert_Calibration contains the YEAR column from Fert_County_1822_long
# plus FERT_kg (and other columns) from Fertilizer_County_Crop


Fert_Calibration <- Fert_Calibration %>%
  rename(COUNTY = County) %>%
  group_by(STATE, COUNTY, YEAR) %>%
  summarize(
    Fert_NIR = mean(farm_val, na.rm = TRUE),
    Fert_Estimated = sum(FERT_kg, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(RATIO = Fert_Estimated/Fert_NIR)


#write.csv(Fert_Calibration,"Outputs/3.6_Fertilizer_Calibration.csv",row.names = FALSE)
```



## Check the crop and state over/underestimated 
This part is not important as we're going to provide implied emission factor. The overestimated or underestimated fertilizer use will not influence the results. 

```{r Estimate check}
library(dplyr)
library(stringr)

library(dplyr)

# fert_calib: Must have columns STATE, County, YEAR, RATIO
# fert_detail: Must have columns STATE, COUNTY, YEAR, CROP, etc.
over_under_crops <- function(fert_calib, fert_detail, threshold = 1) {
  # 1) Subset over & under
  over  <- fert_calib %>% filter(RATIO > threshold)
  under <- fert_calib %>% filter(RATIO < threshold)

  # 2) Join each subset with the crop-level data
  over_with_crops <- over %>%
    left_join(fert_detail, by = c("STATE", "County" = "COUNTY", "YEAR"))

  under_with_crops <- under %>%
    left_join(fert_detail, by = c("STATE", "County" = "COUNTY", "YEAR"))

  # 3) Return as a list
  list(
    over  = over_with_crops,
    under = under_with_crops
  )
}

# Example usage:
results <- over_under_crops(
  fert_calib  = Fert_Calibration,            # your RATIO table
  fert_detail = Fertilizer_County_Crop_clean # your crop-level data
)

# Inspect overestimated counties/crops
head(results$over)

# Inspect underestimated counties/crops
head(results$under)




# 1) Summarize overestimated rows (RATIO > 1) by STATE and CROP
over_by_state_crop <- results$over %>%
  group_by(STATE, CROP) %>%
  summarize(
    total_FERT_kg       = sum(FERT_kg, na.rm = TRUE),
    total_Fert_Estimated= sum(Fert_Estimated, na.rm = TRUE),
    avg_Ratio           = mean(RATIO, na.rm = TRUE), 
    .groups = "drop"
  ) %>%
  arrange(STATE, CROP)

# 2) Summarize underestimated rows (RATIO < 1) by STATE and CROP
under_by_state_crop <- results$under %>%
  group_by(STATE, CROP) %>%
  summarize(
    total_FERT_kg       = sum(FERT_kg, na.rm = TRUE),
    total_Fert_Estimated= sum(Fert_Estimated, na.rm = TRUE),
    avg_Ratio           = mean(RATIO, na.rm = TRUE), 
    .groups = "drop"
  ) %>%
  arrange(STATE, CROP)

# Review the results
head(over_by_state_crop)
head(under_by_state_crop)


```