---
title: "Fertilizer application methods"
author: "Dr. Chih-Yu Hung"
date: "2025-04-28"
output: html_document
---

```{r setup, include=FALSE}
source("setup.R")
```

## To obtain fertilization application method in different regions in the US. 

The CEAP II (2016) have results for incorporation percentage for different regions. 
They group the fertilizer incorporation area into all (100% incorporated), some (X% incorporated), and none (0% incorporated). We will estimate fertilizer incorporation proportion based on CEAP II results. 

Here is the considerations and data we have to estimate the proportion of application method. 
1. Anhydrous ammonia must be incorporated. 
2. Assume Urea, UAN, and other fertilizers share the same incorporation. If one of fertilizer can't deplete incorporation, the unused quota can be evenly allocated to the rest of other fertilizers. 
3. It is possible that my data has exclusive result. for example, only 20% of land is incorporated, but in that state, it uses 60% of anhydrous ammonia. It is because the incorporated land is not crop specific and both the incorporated land and fertilizer types are estimated based on survey. 
4. Pro-rata: Stay neutrality for crops. Every crop places the same fraction of non Anhyrous ammonia (not necessarily incorporated). 
5. Assume application method is identical when double cropping presence.

```{r Read data}
Fertilizer_form <- read.csv("Outputs/3.2_Fertilizer_form.csv")
Fertilizer_use <- read.csv("Outputs/3.4_Fertilizer_Use_County_Crop.csv") %>%
  select(-AREA,-RATE_kg_ha)
CEAP_Fertilization <- read_xlsx("Inputs/Fertilization practice_CEAP.xlsx")
#Agr_area_county<- read.csv("Outputs/3.3_Area_County_Crop.csv")
```


## Prepare data before estimate


```{r data preparatiin}
# 1.  State‑level fertilizer‑form shares
fert_form_clean <- Fertilizer_form %>%
  mutate(
    FORM = str_replace_all(FORM,                         # normalise names
                           c("Urea ammonium nitrate" = "UAN",
                             "Anhydrous ammonia"      = "AA")),
    Fertilizer_ID = recode(FORM,
                           "Urea"  = 1L,
                           "UAN"   = 2L,
                           "AA"    = 3L,
                           "Other" = 4L,
                           .default = NA_integer_)
  ) %>%
  select(STATE, CROP, CLASS, Fertilizer_ID, PERCENTAGE)

# 2.  County‑year totals of N applied by crop class
fert_use_agg <- Fertilizer_use %>%
  select(STATE, COUNTY, YEAR, CROP, CLASS, FERT_kg) %>%
  group_by(STATE, COUNTY, YEAR, CROP, CLASS) %>%           # drop DOUBLE
  summarise(FERT_kg = sum(FERT_kg, na.rm = TRUE), .groups = "drop")

# 3.  CEAP region‑level incorporation fractions (CEAP stratum 2 only)
ceap_frac <- CEAP_Fertilization %>%
  filter(CEAP == 2) %>%                                    # keep stratum 2
  transmute(Region,                                         # two‑letter CEAP code
            inc_type   = Incorporation_type,               # All/Some/None
            inc_share  = Incorporation_percent/100)        # convert to fraction

# 4.  attach CEAP code to every state name 
state2ceap <- state_crosswalk %>%
  select(full, CEAP) %>%
  rename(STATE = full, Region = CEAP)

# 5. ceap_frac_state: per‐state incorporated / some / none shares
ceap_frac_state <- state2ceap %>%
  left_join(ceap_frac, by = "Region") %>%                       # adds inc_type & inc_share
  mutate(STATE = toupper(STATE))


# 6.  CEAP → state: keep All & Some separately 
incorp_state <- ceap_frac_state %>%          # CEAP = 2 only, from earlier step
  summarise(A = sum(inc_share[inc_type == "All" ], na.rm=TRUE),
            S = sum(inc_share[inc_type == "Some"], na.rm=TRUE),
            .by = STATE)                     # A = “all”, S = “some” (0–1 scale)

# 7.  put fertiliser shares (p1 … p4) in wide form 
fert_share_wide <- fert_form_clean %>%                    # still state‑level
  select(STATE, CROP, CLASS, Fertilizer_ID, PERCENTAGE) %>%
  mutate(Fertilizer_ID = factor(Fertilizer_ID, levels = 1:4)) %>%
  pivot_wider(names_from = Fertilizer_ID,
              values_from = PERCENTAGE,
              values_fill = 0,
              names_prefix = "p")        
```

## Calcualte the incorporation proportion
```{r inc calculation}
#  Join with fertiliser shares and apply the new logic -------------
state_crop_class <- fert_share_wide %>%      # p1 … p4 = Urea,UAN,AA,Other
  left_join(incorp_state, by = "STATE") %>%          # add A,S
  replace_na(list(A = 0, S = 0)) %>%
  rowwise() %>%
  mutate(
    # ------------------------------------------------------------------
    # demanded AA incorporation & available capacity
    need_AA   = p3,
    cap_total = A + S,
    # ------------------------------------------------------------------
    # AA takes first from A, then from S
    inc3 = need_AA,                  #              AA gets all it needs
    A_left = ifelse(need_AA >= A, 0, A - need_AA), # "All" left after AA
    S_drawn = pmax(need_AA - A, 0),           # what AA had to draw from “Some”
    S_left  = pmax(S - S_drawn, 0),           # remainder of “Some”
    # ------------------------------------------------------------------
    # what can other ferts still incorporate?
    addl_incorp = A_left + 0.5 * S_left,         # ½ of remaining “Some”
    n_avail     = (p1 > 0) + (p2 > 0) + (p4 > 0),   # how many (1/2/4) present
    inc_other   = addl_incorp / n_avail,           # split evenly to 1,2,4
    
    inc1_init   = ifelse(p1 > 0, pmin(p1, inc_other), 0),
    inc2_init   = ifelse(p2 > 0, pmin(p2, inc_other), 0),
    inc4_init   = ifelse(p4 > 0, pmin(p4, inc_other), 0),
    used_init   = inc1_init + inc2_init + inc4_init,
    rem_incorp  = addl_incorp - used_init,         # still unused capacity
    # cap at each fertiliser’s total share
    inc1 = pmin(p1, inc_other),
    inc2 = pmin(p2, inc_other),
    inc4 = pmin(p4, inc_other),
    
    # remaining head‑room for each fert
    head1   = pmax(p1 - inc1_init, 0),
    head2   = pmax(p2 - inc2_init, 0),
    head4   = pmax(p4 - inc4_init, 0),
    head_sum = head1 + head2 + head4,
    
    # ── 2∘ pass: distribute remainder proportional to head‑room ───────
    incr_factor = ifelse(head_sum > 0, rem_incorp / head_sum, 0),
    
    inc1 = inc1_init + head1 * incr_factor,
    inc2 = inc2_init + head2 * incr_factor,
    inc4 = inc4_init + head4 * incr_factor,
    
    # ------------------------------------------------------------------
    broadcast1 = p1 - inc1,
    broadcast2 = p2 - inc2,
    broadcast3 = p3 - inc3,        # will be zero except if AA > capacity (rare)
    broadcast4 = p4 - inc4,
    .keep = "unused"
  ) %>%
  ungroup() %>%
  rename_with(~ gsub("_", "", .x),              # drop underscores
              starts_with(c("broadcast_", "inc_"))) %>%
  select(-incother)                             # drop incother
```

## Arrange for results and export
```{r export data}
# reshape to long form with Method_ID & Proportion ──────────────
inc_long <- state_crop_class %>%                       # from step2
  select(STATE, CROP, CLASS,
         starts_with("broadcast"), inc1, inc2, inc3, inc4) %>%
  pivot_longer(
    cols           = everything()[-(1:3)],             # all *_1 … *_4
    names_to       = c("type", "Fertilizer_ID"),        #   split name
    names_pattern  = "(broadcast|inc)_?(\\d)",            #   by regex
    values_to      = "share_abs"
  ) %>%
  mutate(
    Fertilizer_ID = as.integer(Fertilizer_ID),          # "1","2",… → int
    Method_ID     = if_else(type == "broadcast", 1L, 2L)
  ) %>%
  ## convert absolute share to proportion within each fertiliser
  group_by(STATE, CROP, CLASS, Fertilizer_ID) %>%
  mutate(
  total_share = sum(share_abs, na.rm = TRUE)) %>%
  mutate(Proportion = ifelse(total_share >0, share_abs / total_share, 0)) %>%
  ungroup() %>%
  select(STATE, CROP, CLASS, Fertilizer_ID, Method_ID, Proportion) %>%
  arrange(STATE, CROP, CLASS, Fertilizer_ID, Method_ID)

## final_table  contains:
## STATE | CROP | CLASS | Fertilizer_ID | Method_ID | Proportion
#write.csv(inc_long,"Outputs/3.7 Fertilizer app method_STATE.csv", row.names = FALSE)


# expand to county × year, not necessary for STATE level data
# final_table <- fert_use_agg %>%                  # keys STATE,COUNTY,YEAR,CROP,CLASS
#   distinct(STATE, COUNTY, YEAR, CROP, CLASS) %>%
#   left_join(inc_long, by = c("STATE", "CROP", "CLASS")) %>%
#   arrange(STATE, COUNTY, YEAR, CROP, CLASS, Fertilizer_ID, Method_ID) %>%
#   filter(Proportion >= 0)

## final_table  contains:
## STATE | COUNTY | YEAR | CROP | CLASS | Fertilizer_ID | Method_ID | Proportion
#write.csv(final_table,"Outputs/3.7 Fertilizer app method_STATE.csv", row.names = FALSE)



```