---
title: "US Inorganic fertilization"
author: "Dr. Chih-Yu Hung"
date: "2025-01-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(data.table)
```

## US Inorganic fertilization

This file shows how we estimate inorganic fertilization used in the US. This is primairly based on USA's 2024 NIR "Annex 3.12 Methodologies for SOC, N2O and CH4 from agricultural lands." 

USA's 2024 NIR showed their national synthetic N fertilzation to cropland is 11305, 11249, and 11209 kt N in 2018, 2019, and 2020. (Table A-181 in Annex 3.12)


Ref
Brakebill, J.W. and Gronberg, J.M., 2017, County-Level Estimates of Nitrogen and Phosphorus from Commercial Fertilizer for the Conterminous United States, 1987-2012: U.S. Geological Survey data release, https://doi.org/10.5066/F7H41PKX.

```{r Fertilization County level}
#Data from Brakebill and Gronberg 2017. (1987-2012 only) 
#According to USA CRT 2012, Inorganic N fertilizer is 11,918,834,756 kg N (NOT t N, the unit is wrong in their CRT), it's equal to sum of farm N 2012 in Brakebill's file
Fert_County_2012 <- read.csv("Fertilizer/CNTY_FERT_1987-2012.csv",header=T) %>%
                   select(State, County, farmN2012)    

#Assuming there is no change of the proportion between counties. 
#We are going to extrapolate fertilizer use based on CRT values. 
USCRT <- "C:/Users/HungC/OneDrive - EC-EC/2023/Fuel LCA/USA NIR/US CRT/USA-CRT-2024-V0.5-submission-started/"

# Generate the pattern for file matching
file_pattern <- "USA-CRT-2024-V0.5-(201[2-9]|202[0-2])"

# List all files in the folder that match the pattern
excel_files <- list.files(path = USCRT, pattern = paste0(file_pattern, ".*\\.xlsx$"), full.names = TRUE)

# Extract year from file name using regex
extract_year <- function(file_name) {
  sub(".*V0.5-(\\d{4}).*", "\\1", file_name)
}

# Function to read and extract the desired value
extract_value <- function(file_path) {
  tryCatch({
    # Read the specific cell from the sheet "Table3.D"
    value <- read_excel(file_path, sheet = "Table3.D", range = "D11", col_names = FALSE, .name_repair = "minimal")
    as.numeric(value[[1]])  # Return the numeric value
  }, error = function(e) {
    NA  # Return NA if an error occurs
  })
}

# Create the table
results <- data.frame(
  Year = sapply(excel_files, extract_year),
  `Inorganic N fertilizer` = sapply(excel_files, extract_value)
)

# Convert the Year column to numeric for proper sorting
results <- results %>% arrange(as.numeric(Year))

#Ratio based on 2012 fertilizer. This ratio is applying to fertilizer at county level in 2012. 
results$Ratio <- results$Inorganic.N.fertilizer/results$Inorganic.N.fertilizer[1]

#The fertilization in 2018-2022
Fert_County_1822 <- Fert_County_2012 %>%
  mutate(farmN2018 = farmN2012 *results$Ratio[results$Year == 2018]) %>%
  mutate(farmN2019 = farmN2012 *results$Ratio[results$Year == 2019]) %>%
  mutate(farmN2020 = farmN2012 *results$Ratio[results$Year == 2020]) %>%
  mutate(farmN2021 = farmN2012 *results$Ratio[results$Year == 2021]) %>%
  mutate(farmN2022 = farmN2012 *results$Ratio[results$Year == 2022]) 


```


### BEfore I do anything remember that this fertilzation amount is the total of all crops. I only want some crops. The sum of fertilzer of the interested crop must less than the total fertilizer in this list. 
(CY, 2025 Jan 09)

```{r retrieve survey data for fertilzation rate}
#  US_environment <- fread("qs.environmental_20250114.txt/qs.environmental_20250114.txt")
# # 
# Fert_rate <- US_environment %>%
#   filter(SOURCE_DESC == "SURVEY") %>% #keep CENSUS data only
#   filter(!(STATE_ALPHA %in% c("AK","GU","AS","PR","VI","MP","","HI",""))) %>% #exclude territories, HI and AK
#   filter(YEAR %in% c(2018:2022)) %>%
#   filter(COMMODITY_DESC %in% c("CORN","SORGHUM","GRAIN",
#                                "WHEAT","SOYBEANS","OATS","BARLEY",
#                                "SWEET CONR","PEAS","RYE","CANOLA",
#                                "TRITICALE","CAMELINA")) %>%
#   filter(DOMAINCAT_DESC == "FERTILIZER: (NITROGEN)") %>%
#   filter(UNIT_DESC == "LB / ACRE / YEAR, AVG") %>%
#   select(COMMODITY_DESC,CLASS_DESC,STATE_ALPHA,STATE_NAME,YEAR,VALUE)
# save(Fert_rate, file = "Inputs/Fertrate_2018-22.RData")

load("Inputs/Fertrate_2018-22.RData")

Fert_rate <- Fert_rate %>%
  filter(VALUE != "(D)") %>%
  mutate(VALUE = as.numeric(VALUE)) %>%
  mutate(CROP = paste(COMMODITY_DESC, CLASS_DESC, sep ="_")) %>%
  group_by(CROP,STATE_NAME)%>%
  summarise(RATE = mean(VALUE))

```

## Complete the data table

There are crops and states missing because they are not included in the survey. 
Crop: camelina, canola, rye_fall, rye_winter, triticale


```{r complete data table}
missing_crops <- c("CAMELINA", "CANOLA", "RYE_FALL", "RYE_WINTER", "TRITICALE")

states <- toupper(c(
  "Alabama", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", 
  "Florida", "Georgia", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", 
  "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", 
  "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", 
  "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", 
  "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", 
  "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", 
  "Wisconsin", "Wyoming"
))

# Create the full combination of crops and states
expanded_data <- expand_grid(CROP = c(unique(Fert_rate$CROP), missing_crops), STATE_NAME = states)

# Join the original data with the expanded grid
Fert_rate_expanded <- expanded_data %>%
  left_join(Fert_rate, by = c("CROP", "STATE_NAME"))


```


## USing the USDA Farm Production Regions

USDA Farm production regions are useful to estimate fertilizer rate where is missing. Each region defines a geographic area in which crops and cropping practices are similiar. (Cooter et al., 2012)

Cooter et al., 2012. doi:10.5194/bg-9-4023-2012



